Exploring dimensions of biodiversity in Japanese ferns

### 2021-05-28 ms/manuscript.Rmd
Summarize model stats

```{r setup, include = FALSE}
# Load libraries used for rendering Rmds
library(officedown)
library(officer)
library(flextable)
library(kableExtra)

@@

```{r load-data}
# Load objects from targets workflow
tar_load(c(
  "occ_point_data_ferns",
  "occ_point_data_ferns_unfiltered",
  "biodiv_ferns_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "signif_cells_protected_area",
  "protected_areas",
  "japan_shp",
  "phy_sig_results",
  "binary_sig_results",
  "traits_summary",
  "fern_traits",
  "japan_pterido_tree",
  "japan_fern_tree",
  "ppgi",
  "occ_point_data_ferns",
  "comm_ferns_full",
  "comm_ferns",
  "japan_points_raw",
  # "plastome_calibration_dates",
  # FIXME: remove until phylo analysis is working "plastome_tree",
  "green_list"
))
```

@@

We first tested for spatial autocorrelation in biodiversity metrics and independent variables (environmental variables and % apomictic taxa) using Moran's I with the `moran.mc` function in the `spdep` package. In all cases, significant spatial autocorrelation was detected (Moran's I > 0, *P* value of permutation test < 0.05). Therefore, we used spatially explicit methods for modeling.

We checked for correlation between independent variables while taking into account spatial autocorrelation using a modified *t*-test with the `modified.ttest` in the `SpatialPack` R package. This indicated that temperature seasonality and % apomictic taxa were correlated with mean annual temperature. The dataset including % apomictic taxa is smaller than the full dataset (_ vs _ taxa, respectively), so for the main modeling analysis we only included the following environmental variables: temperature, precipitation, and precipitation seasonality.

We constructed a linear model for each response variable (richness, SES of PD, SES of FD, SES of RPD, SES of RFD, PE p-score) dependent on the environmental variables. We did not include any interaction terms as we had no a-priori reason to do so. Initial inspection of the relationship between richness and temperature showed a hump-shaped pattern, so we also included a quadratic term for temperature for the richness model only. We fit models including a Matérn correlation matrix based on the geographic centroids of each grid cell as a random effect with the R package spaMM [@Rousset2014]. We verified the significance of each term of the model by conducting likelihood ratio tests (LRTs) between the full model and models each with a single term of interest removed, in sequence. We also verified the significance of the full model relative to the null model (only including the Matérn correlation matrix) using LRTs.

We postulated that % apomictic taxa might explain the distribution of genetic diversity better than temperature, so we also constructed a limited set of models to test this hypothesis. These models included response variables with a genetic component (SES of PD, SES of RPD, PE p-score) as dependent on either % apomictic taxa or temperature, precipitation, and precipitation seasonality, with random spatial effects as described above. We compared the fit between models including % apomictic taxa or temperature using conditional Akaike Information Criterion (cAIC).

@@

```{r phy-sig-stats}
# Verify that min lambda >0.95
# assert_that(phy_sig_results %>% pull(lambda) %>% min > 0.95)
# Verify that max K < 0.0.6
# assert_that(phy_sig_results %>% pull(kval) %>% max < 0.06)

@@

<!-- All continuous traits showed significant phylogenetic signal, but estimates of signal strength differed between *K* and lambda: all lambda values were > 0.95, indicating evolution by BM; however, values of *K* were < 0.06, indicating less phylogenetic signal than expected under BM (`r s_table("phy-sig")`). `r n_negative_d` out of `r nrow(binary_sig_results)` of continuous traits (coded as binary) showed as much or more phylogenetic signal as expected under BM (*D* =< 0); others had *D* > 0 (`r s_table("phy-sig-binary")`). -->

@@

### Correlations between biodiversity and reproductive mode

Spatial autocorrelation as measured with Moran's I was detected in linear models, but was not in any of the spatial mixed linear models. SES of RPD in particular shows a strong negative correlation with % apomictic taxa (_ estimated effect _, SD).

@@

@@

>>>>

```{r setup, include = FALSE}
# Load libraries used for rendering Rmds
library(officedown)
library(officer)
library(flextable)
library(kableExtra)
library(ftExtra)

@@

```{r load-data}
# Load objects from targets workflow
tar_load(c(
  "occ_point_data_ferns",
  "occ_point_data_ferns_unfiltered",
  "biodiv_ferns_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "signif_cells_protected_area",
  "protected_areas",
  "japan_shp",
  "phy_sig_results",
  "binary_sig_results",
  "traits_summary",
  "fern_traits",
  "japan_pterido_tree",
  "japan_fern_tree",
  "ppgi",
  "occ_point_data_ferns",
  "comm_ferns_full",
  "comm_ferns",
  "japan_points_raw",
  # "plastome_calibration_dates",
  # FIXME: remove until phylo analysis is working "plastome_tree",
  "green_list",
  "env_model_params",
  "repro_model_params",
  "aic_env_repro"
))
```

@@

We first tested for spatial autocorrelation in biodiversity metrics (richness, SES of PD, SES of FD, SES of RPD, SES of RFD, PE p-score) and independent variables (environmental variables and % apomictic taxa) using Moran's I with the `r func("moran.mc")` function in the `r pack("spdep")` package. In all cases, significant spatial autocorrelation was detected (Moran's I > 0, *P* value of permutation test < 0.05). Therefore, we used spatially explicit methods for modeling.

We checked for correlation between independent variables while taking into account spatial autocorrelation using a modified *t*-test with the `r func("modified.ttest")` function in the `r pack("SpatialPack")` R package. This indicated that temperature seasonality and % apomictic taxa were correlated with mean annual temperature. The dataset including % apomictic taxa is smaller than the full dataset (_ vs _ taxa, respectively), so for the main modeling analysis we only included the following environmental variables: temperature, precipitation, and precipitation seasonality.

We constructed a linear model for each biodiversity metric dependent on the environmental variables. Initial inspection of the relationship between richness and temperature showed a hump-shaped pattern, so we also included a quadratic term for temperature for the richness model only. We fit models including a Matérn correlation matrix based on the geographic centroids of each grid cell as a random effect with the R package `r pack("spaMM")` [@Rousset2014]. We verified the significance of each term of the model by conducting likelihood ratio tests (LRTs) between the full model and models each with a single term of interest removed, in sequence. We also verified the significance of the full model relative to the null model (only including the Matérn correlation matrix) using LRTs.

We postulated that % apomictic taxa might explain the distribution of genetic diversity better than temperature, so we also constructed a limited set of models to test this hypothesis. These models included biodiversity metrics with a genetic component (SES of PD, SES of RPD, PE p-score) as dependent on either % apomictic taxa or temperature, precipitation, and precipitation seasonality, with random spatial effects as described above. We compared the fit between models including % apomictic taxa or temperature using conditional Akaike Information Criterion (cAIC) [@Courtiol2017; @Vaida2005].

@@

```{r phy-sig-stats}
# Verify that min lambda >0.95
assert_that(phy_sig_results %>% pull(lambda) %>% min > 0.95)
# Verify that max K < 0.0.6
assert_that(phy_sig_results %>% pull(kval) %>% max < 0.06)

@@

All continuous traits showed significant phylogenetic signal, but estimates of signal strength differed between *K* and lambda: all lambda values were > 0.95, indicating evolution by BM; however, values of *K* were < 0.06, indicating less phylogenetic signal than expected under BM (`r s_table("phy-sig")`). `r n_negative_d` out of `r nrow(binary_sig_results)` of continuous traits (coded as binary) showed as much or more phylogenetic signal as expected under BM (*D* =< 0); others had *D* > 0 (`r s_table("phy-sig-binary")`).

@@

### Relationships between biodiversity and climate or reproductive mode

All biodiversity metrics except for SES of RFD showed a significant relationship with temperature (`r table("env-model-params")`).

@@

```{r table-prep}
# Define nicely formatted terms for printing variable names
resp_var_lookup <- tribble(
  ~resp_var, ~resp_var_pretty,
  "fd_obs_z", "SES of FD",
  "pd_obs_z", "SES of PD",
  "pe_obs_p_upper", "PE *p*-value",
  "rfd_obs_z", "SES of RFD",
  "richness", "Richness",
  "rpd_obs_z", "SES of RPD",
)

indep_var_lookup <- tribble(
  ~term, ~term_pretty,
  "temp", "Temperature",
  "precip", "Precipitation",
  "precip_season", "Precipitation seasonality",
  "I(temp^2)", "Temperature^2^",
  "percent_apo", "% apomictic taxa"
)
```

@@

```{r env-model-params-prep}
# Format model parameter summary for nice printing as a table
env_model_params_pretty <-
env_model_params %>%
  # Remove intercept
  filter(term != "(Intercept)") %>%
  # Double check that log-likelihood of null model is less than full
  verify(all(loglik_null < loglik_full)) %>%
  select(resp_var, term, estimate, cond_se, t_value, p_value) %>%
  # Format response variable and fixed effect names to look nice
  left_join(resp_var_lookup, by = "resp_var") %>%
  select(-resp_var) %>%
  rename(resp_var = resp_var_pretty) %>%
  left_join(indep_var_lookup, by = "term") %>%
  select(-term) %>%
  rename(term = term_pretty) %>%
  # Round to 2 signif digits
  mutate(across(c(estimate, cond_se, t_value), ~signif(., 2))) %>%
  # Format p-value
  mutate(
    signif = if_else(p_value < 0.05, TRUE, FALSE),
    p_value = case_when(
      p_value < 0.0001 ~ "< 0.0001",
      TRUE ~ signif(p_value, 2) %>% as.character
    )
  ) %>%
  arrange(resp_var, term) %>%
  select(resp_var, term, estimate, cond_se, t_value, p_value, signif)
```

```{r env-model-params-doc, results = "asis", eval = params$doc_type == "doc"}
env_model_params_pretty %>%
  mutate(
    across(where(is.numeric), as.character)) %>%
  rename("*t*-value" = t_value, "*p*-value" = p_value, "Variable" = term, "Estimate" = estimate, "Conditional SE" = cond_se) %>%
  as_grouped_data(groups = "resp_var") %>%
  as_flextable(hide_grouplabel = TRUE, col_keys = c("Variable", "Estimate", "Conditional SE", "*t*-value", "*p*-value")) %>%
  bold(i = ~ signif == TRUE, bold = TRUE) %>%
  colformat_md(part = "header")
```

```{r env-model-params-pdf, results = "asis", eval = params$doc_type == "pdf"}
# Define row groups
kbl_groups <-
env_model_params_pretty %>%
  transmute(resp_var, row_num = 1:nrow(.)) %>%
  group_by(resp_var) %>%
  summarize(start = min(row_num), end = max(row_num), .groups = "drop") %>%
  split(.$resp_var)

signif_rows <- env_model_params_pretty %>% pull(signif) %>% which()

env_model_params_pretty %>%
  mutate(
    across(where(is.numeric), as.character)) %>%
  rename("*t*-value" = t_value, "*p*-value" = p_value, "Variable" = term, "Estimate" = estimate, "Conditional SE" = cond_se) %>%
  select(-resp_var, -signif) %>%
  kbl(booktabs = TRUE) %>%
  kable_styling() %>%
  pack_rows(kbl_groups[[1]]$resp_var, kbl_groups[[1]]$start, kbl_groups[[1]]$end) %>%
  pack_rows(kbl_groups[[2]]$resp_var, kbl_groups[[2]]$start, kbl_groups[[2]]$end) %>%
  pack_rows(kbl_groups[[3]]$resp_var, kbl_groups[[3]]$start, kbl_groups[[3]]$end) %>%
  pack_rows(kbl_groups[[4]]$resp_var, kbl_groups[[4]]$start, kbl_groups[[4]]$end) %>%
  pack_rows(kbl_groups[[5]]$resp_var, kbl_groups[[5]]$start, kbl_groups[[5]]$end) %>%
  pack_rows(kbl_groups[[6]]$resp_var, kbl_groups[[6]]$start, kbl_groups[[6]]$end) %>%
  row_spec(row = signif_rows, bold = TRUE)
```

**`r table("env-model-params")`**. Parameter estimates of the effect of environmental variables on biodiversity metrics in the ferns of Japan. Significant effects (*p* < 0.05) in bold; significance assessed with likelihood ratio tests (see Methods). All models include random effect of geographical coordinates.
